---
title: "Actual vs. Statutory Effective Tariff Rates, 2025--2026"
author: "The Budget Lab at Yale"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
  word_document:
    toc: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)

here::i_am("R/etr_eval_report.Rmd")
source(here("R", "utils.R"))
theme_set(theme_etr())

# ==============================================================================
# DATA LOADING
# ==============================================================================

# Actual ETR (customs duties / imports, monthly from Census/Treasury via Haver)
actual_etr <- load_actual_etr() %>%
  filter(date >= "2024-01-01") %>%
  mutate(actual_rate = effective_rate / 100)

# Daily statutory ETR (import-weighted, from tariff-rate-tracker)
daily_statutory <- load_daily_etr() %>%
  filter(date <= Sys.Date())

# Authority decomposition
daily_auth <- load_daily_authority()

# Monthly statutory (first-of-month snapshot)
monthly_statutory <- collapse_statutory_monthly(daily_statutory)

# Revision dates for policy event context
revision_dates <- load_revision_dates()

# Census HS2 x country x month (from 01_pull_census_trade.R)
census_file <- here("data", "census_hs2_country_monthly.csv")
census_available <- file.exists(census_file)
if (census_available) {
  census <- load_census_trade() %>%
    mutate(partner_group = assign_partner_group(cty_code))
}

# Statutory timeseries (product-level)
ts_file <- file.path(TRACKER_DIR, "data", "timeseries", "rate_timeseries.rds")
ts_available <- file.exists(ts_file)
if (ts_available) {
  ts <- load_timeseries()
  latest_date <- min(Sys.Date(), max(ts$valid_until, na.rm = TRUE))
  current_rates <- get_rates_at_date(ts, latest_date)
  baseline_rates <- get_rates_at_date(ts, "2025-01-01")

  chapter_summary <- current_rates %>%
    mutate(hs2 = substr(hts10, 1, 2)) %>%
    group_by(hs2) %>%
    summarize(
      n_products = n_distinct(hts10),
      mean_total = mean(total_rate, na.rm = TRUE),
      mean_232 = mean(rate_232, na.rm = TRUE),
      mean_301 = mean(rate_301, na.rm = TRUE),
      mean_ieepa = mean(rate_ieepa_recip, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_total))

  baseline_by_ch <- baseline_rates %>%
    mutate(hs2 = substr(hts10, 1, 2)) %>%
    group_by(hs2) %>%
    summarize(baseline_rate = mean(total_rate, na.rm = TRUE), .groups = "drop")
}

# ==============================================================================
# FIGURE/TABLE METADATA
# ==============================================================================

FIGURE_METADATA <- list(
  F1 = list(
    label    = "Figure 1",
    title    = "Actual vs. Statutory Effective Tariff Rate",
    subtitle = "Monthly actual (customs duties / imports) vs. daily statutory (import-weighted)",
    source   = "U.S. Treasury/Census via Haver Analytics; The Budget Lab Tariff Rate Tracker",
    notes    = "Actual rate is customs duties as a share of goods import value (monthly). Statutory rate is the import-weighted average of all HTS tariff schedules by calendar day.",
    y_label  = "Percent"
  ),
  F2 = list(
    label    = "Figure 2",
    title    = "Statutory-Actual ETR Gap",
    subtitle = "Percentage point difference (statutory minus actual) by month",
    source   = "The Budget Lab analysis",
    notes    = "Positive values indicate the statutory rate exceeds what is actually collected.",
    y_label  = "Percentage Points"
  ),
  F3 = list(
    label    = "Figure 3",
    title    = "Monthly Import Volumes by Trading Partner",
    subtitle = "Billions USD (consumption value), 2024--2025",
    source   = "Census Bureau International Trade API",
    notes    = "",
    y_label  = "Billions USD"
  ),
  F4 = list(
    label    = "Figure 4",
    title    = "Import Surge by HS Chapter: 2025 vs. 2024",
    subtitle = "Year-over-year import growth (%) in Q1 2025 vs. tariff increase (pp)",
    source   = "Census Bureau International Trade API; The Budget Lab Tariff Rate Tracker",
    notes    = "Each point is an HS2 chapter. Chapters in the upper-right experienced both large tariff increases and large import surges, consistent with front-running.",
    y_label  = "Q1 2025 YoY Import Growth (%)"
  ),
  F5 = list(
    label    = "Figure 5",
    title    = "Transshipment Indicator: Import Shifts by Origin",
    subtitle = "Change in import share from China vs. Vietnam/Thailand/Malaysia/Cambodia",
    source   = "Census Bureau International Trade API",
    notes    = "A negative correlation suggests trade diversion through transshipment countries.",
    y_label  = ""
  ),
  F6 = list(
    label    = "Figure 6",
    title    = "Dutiable Share by HS Chapter",
    subtitle = "Top 10 chapters by 2025 import value; dutiable value / consumption value",
    source   = "Census Bureau International Trade API",
    notes    = "Dutiable value is the portion of imports subject to duty, not the duty amount itself. High shares indicate chapters where most trade faces tariffs.",
    y_label  = "Percent"
  ),
  F7 = list(
    label    = "Figure 7",
    title    = "Tariff Gap Decomposition: Potential Explanatory Factors",
    subtitle = "Estimated contribution of avoidance channels to the statutory-actual gap",
    source   = "The Budget Lab analysis",
    notes    = "Decomposition is illustrative; channels are not fully separable from aggregate data alone.",
    y_label  = "Percentage Points"
  )
)

fig_meta <- function(id, field) {
  val <- FIGURE_METADATA[[id]][[field]]
  if (is.null(val)) "" else val
}
fig_caption <- function(id) {
  src <- fig_meta(id, "source")
  notes <- fig_meta(id, "notes")
  cap <- paste0("Source: ", src)
  if (nzchar(notes)) cap <- paste0(cap, "\nNote: ", notes)
  cap
}

# ==============================================================================
# COMPUTED OBJECTS
# ==============================================================================

# Merge actual and statutory at monthly level
comparison <- monthly_statutory %>%
  select(date, statutory_rate = weighted_etr) %>%
  full_join(
    actual_etr %>% select(date, actual_rate),
    by = "date"
  ) %>%
  filter(date >= "2025-01-01") %>%
  arrange(date) %>%
  mutate(
    gap_pp = (statutory_rate - actual_rate) * 100,
    statutory_pct = statutory_rate * 100,
    actual_pct = actual_rate * 100
  )

latest_comp <- comparison %>% filter(!is.na(actual_rate)) %>% filter(date == max(date))
latest_month <- format(latest_comp$date, "%B %Y")
latest_statutory <- sprintf("%.1f", latest_comp$statutory_pct)
latest_actual <- sprintf("%.1f", latest_comp$actual_pct)
latest_gap <- sprintf("%.1f", latest_comp$gap_pp)
```

## Summary

This report compares two measures of tariff rates on U.S. imports during the 2025--2026 tariff escalation:

- **Statutory ETR**: The import-weighted average tariff rate implied by the Harmonized Tariff Schedule, computed from The Budget Lab's [Tariff Rate Tracker](https://budgetlab.yale.edu/research/state-tariffs). This represents what importers *should* pay given the tariff schedule in effect.
- **Actual ETR**: Customs duties actually collected divided by the value of goods imports, from Treasury/Census data via Haver Analytics. This represents what importers *actually* paid.

As of `r latest_month`, the statutory ETR was `r latest_statutory`% while the actual collected rate was `r latest_actual`%, a gap of `r latest_gap` percentage points. This gap has several potential explanations---timing lags between tariff announcements and collection, preferential trade agreement utilization (USMCA), tariff avoidance through supply chain restructuring, stockpiling ahead of tariff effective dates, and outright evasion---which we explore below.

---

## Figure 1: Actual vs. Statutory ETR

The statutory rate (navy line) represents the daily import-weighted average tariff rate from the HTS schedule. The actual rate (red dots) shows monthly customs duties as a share of import value. The actual rate is inherently lagged: tariffs announced in month $t$ may not appear in collection data until months $t+1$ or $t+2$ due to processing times, bonding, and liquidation schedules.

```{r fig-actual-vs-statutory}
ggplot() +
  geom_step(data = daily_statutory,
            aes(x = date, y = weighted_etr * 100),
            color = color_statutory, linewidth = 0.7) +
  geom_point(data = actual_etr %>% filter(date >= "2025-01-01"),
             aes(x = date, y = actual_rate * 100),
             color = color_actual, size = 3, shape = 16) +
  geom_vline(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
             aes(xintercept = date), linetype = "dashed",
             color = tbl_gray, linewidth = 0.3) +
  geom_text(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
            aes(x = date, y = Inf, label = label),
            vjust = 1.3, hjust = -0.05, size = 2.3, color = tbl_gray) +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.95,
           label = "Statutory (daily)", color = color_statutory, hjust = 0, fontface = "bold") +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.88,
           label = "Actual (monthly)", color = color_actual, hjust = 0, fontface = "bold") +
  labs(
    title = fig_meta("F1", "title"),
    subtitle = fig_meta("F1", "subtitle"),
    x = NULL, y = fig_meta("F1", "y_label"),
    caption = fig_caption("F1")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_cartesian(ylim = c(0, NA))
```

---

## Figure 2: The Statutory-Actual Gap

```{r fig-gap}
gap_data <- comparison %>% filter(!is.na(gap_pp))

gap_data %>%
  ggplot(aes(x = date, y = gap_pp)) +
  geom_col(fill = color_gap, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  geom_text(aes(label = sprintf("%.1f pp", gap_pp)),
            vjust = ifelse(gap_data$gap_pp >= 0, -0.5, 1.5),
            size = 3) +
  labs(
    title = fig_meta("F2", "title"),
    subtitle = fig_meta("F2", "subtitle"),
    x = NULL, y = fig_meta("F2", "y_label"),
    caption = fig_caption("F2")
  )
```

The gap between statutory and actual rates is expected to be positive for several structural reasons:

1. **Collection timing**: Tariffs are assessed at entry but may be collected weeks or months later.
2. **USMCA utilization**: Our statutory model assumes 100% utilization for eligible products; actual utilization is ~50%.
3. **Duty drawback and refunds**: Some collected duties are later refunded.
4. **Preferential programs**: GSP, FTAs, and other preference programs reduce effective rates.
5. **De minimis threshold**: Shipments below $800 enter duty-free.

---

## Collapsing Statutory Rates to Monthly

We collapse the daily statutory ETR to monthly using **first-of-month snapshots**. Tariff rates are step functions that change only at discrete policy revision dates. Averaging would create artificial intermediate values. The first-of-month snapshot aligns with Census monthly reporting and is easy to interpret.

```{r tbl-monthly-comparison}
comparison %>%
  filter(!is.na(actual_rate)) %>%
  mutate(
    Month = format(date, "%b %Y"),
    `Statutory ETR (%)` = sprintf("%.2f", statutory_pct),
    `Actual ETR (%)` = sprintf("%.2f", actual_pct),
    `Gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Month, `Statutory ETR (%)`, `Actual ETR (%)`, `Gap (pp)`) %>%
  kable(caption = "Monthly Statutory vs. Actual Effective Tariff Rates",
        align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Import Volumes by Trading Partner

```{r fig-partner-imports, eval=census_available}
# Aggregate Census data to monthly by partner group
partner_monthly <- census %>%
  group_by(date, partner_group) %>%
  summarize(
    imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9,
    .groups = "drop"
  ) %>%
  filter(date >= "2024-01-01", date <= Sys.Date())

# Focus on major partners
major <- c("China", "Canada", "Mexico", "EU", "Japan", "S. Korea", "UK")
partner_major <- partner_monthly %>%
  filter(partner_group %in% major)

partner_major %>%
  ggplot(aes(x = date, y = imports_B, color = partner_group)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = PARTNER_COLORS) +
  labs(
    title = fig_meta("F3", "title"),
    subtitle = fig_meta("F3", "subtitle"),
    x = NULL, y = fig_meta("F3", "y_label"),
    color = NULL,
    caption = fig_caption("F3")
  )
```

```{r fig-partner-imports-na, eval=!census_available, results='asis'}
cat("*Census trade data not available. Run R/01_pull_census_trade.R first.*")
```

---

## Stockpiling Analysis

### Import Surge vs. Tariff Increase by Chapter

The stockpiling hypothesis predicts that chapters facing larger tariff increases should show larger pre-tariff import surges. We test this by comparing Q1 2025 year-over-year import growth against the statutory tariff increase from baseline to current policy.

```{r fig-stockpiling-scatter, eval=census_available && ts_available, fig.height=7}
# Compute Q1 2025 vs Q1 2024 import growth by chapter
q1_2025 <- census %>%
  filter(year == 2025, month %in% 1:3) %>%
  group_by(hs2) %>%
  summarize(imports_q1_2025 = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

q1_2024 <- census %>%
  filter(year == 2024, month %in% 1:3) %>%
  group_by(hs2) %>%
  summarize(imports_q1_2024 = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

import_growth <- q1_2025 %>%
  inner_join(q1_2024, by = "hs2") %>%
  mutate(yoy_growth = (imports_q1_2025 / imports_q1_2024 - 1) * 100) %>%
  filter(is.finite(yoy_growth))

# Merge with tariff increase
rate_change <- chapter_summary %>%
  left_join(baseline_by_ch, by = "hs2") %>%
  mutate(rate_increase_pp = (mean_total - baseline_rate) * 100)

stockpile_data <- import_growth %>%
  inner_join(rate_change %>% select(hs2, rate_increase_pp, n_products), by = "hs2") %>%
  mutate(imports_B = imports_q1_2025 / 1e9)

# Scatter: tariff increase vs import surge
p_scatter <- stockpile_data %>%
  filter(rate_increase_pp > 0.5) %>%
  ggplot(aes(x = rate_increase_pp, y = yoy_growth, size = imports_B)) +
  geom_point(color = color_statutory, alpha = 0.6) +
  geom_text(aes(label = hs2), hjust = -0.3, vjust = 0.5, size = 2.5, color = tbl_gray) +
  geom_hline(yintercept = 0, linetype = "dashed", color = tbl_gray) +
  geom_smooth(method = "lm", se = TRUE, color = color_actual, linewidth = 0.8) +
  scale_size_continuous(range = c(1, 8), labels = dollar_format(prefix = "$", suffix = "B")) +
  labs(
    title = fig_meta("F4", "title"),
    subtitle = fig_meta("F4", "subtitle"),
    x = "Tariff Increase from Jan 2025 Baseline (pp)",
    y = fig_meta("F4", "y_label"),
    size = "Q1 2025 Imports",
    caption = fig_caption("F4")
  )

# Top 15 chapters by import growth, conditional on tariff increase
p_bar <- stockpile_data %>%
  filter(rate_increase_pp > 2) %>%
  arrange(desc(yoy_growth)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(hs2, yoy_growth), y = yoy_growth)) +
  geom_col(fill = color_statutory, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  coord_flip() +
  labs(
    title = "Top 15 Chapters: Q1 Import Growth (Tariff-Affected Only)",
    subtitle = "Among chapters with > 2pp tariff increase",
    x = "HS Chapter", y = "YoY Import Growth (%)"
  )

p_scatter / p_bar +
  plot_annotation(title = "Stockpiling Evidence: Import Surges vs. Tariff Increases")
```

```{r stockpiling-na, eval=!census_available || !ts_available, results='asis'}
cat("*Census trade data or statutory timeseries not available.*")
```

### Monthly Import Dynamics Around Liberation Day

```{r fig-liberation-day, eval=census_available, fig.height=6}
# Focus on chapters with large IEEPA increases (Liberation Day = April 2)
# Show monthly import trajectory for top affected chapters
if (ts_available) {
  ieepa_chapters <- chapter_summary %>%
    filter(mean_ieepa > 0.05) %>%  # Chapters with > 5% IEEPA rate
    arrange(desc(mean_ieepa)) %>%
    head(8) %>%
    pull(hs2)
} else {
  ieepa_chapters <- c("84", "85", "87", "39", "94", "73", "40", "90")
}

census %>%
  filter(hs2 %in% ieepa_chapters, year %in% c(2024, 2025)) %>%
  group_by(date, hs2) %>%
  summarize(imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9, .groups = "drop") %>%
  ggplot(aes(x = date, y = imports_B, color = hs2)) +
  geom_line(linewidth = 0.6) +
  geom_vline(xintercept = as.Date("2025-04-02"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2025-04-02"), y = Inf, label = "Liberation Day",
           vjust = 1.5, hjust = -0.1, size = 3, color = "red") +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  labs(
    title = "Monthly Imports for IEEPA-Affected Chapters",
    subtitle = "Top chapters by IEEPA reciprocal rate; note pre-April surge and post-April dip",
    x = NULL, y = "Billions USD", color = "HS Chapter",
    caption = "Source: Census Bureau International Trade API"
  )
```

---

## Transshipment Analysis

```{r fig-transshipment, eval=census_available, fig.height=7}
# Compare China import decline vs SE Asian import growth
# Census codes: China=5700, Vietnam=5520, Thailand=5490, Malaysia=5570, Cambodia=5550
transship_ctys <- c("5700", "5520", "5490", "5570", "5550")
transship_names <- c("5700" = "China", "5520" = "Vietnam", "5490" = "Thailand",
                     "5570" = "Malaysia", "5550" = "Cambodia")

transship_data <- census %>%
  filter(cty_code %in% transship_ctys) %>%
  mutate(country = transship_names[cty_code]) %>%
  group_by(date, country) %>%
  summarize(imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9, .groups = "drop") %>%
  filter(date >= "2024-01-01", date <= Sys.Date())

# Compute China share vs SE Asian share over time
total_by_month <- census %>%
  group_by(date) %>%
  summarize(total_imports = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

share_data <- census %>%
  filter(cty_code %in% transship_ctys) %>%
  mutate(country = transship_names[cty_code]) %>%
  group_by(date, country) %>%
  summarize(imports = sum(con_val_mo, na.rm = TRUE), .groups = "drop") %>%
  left_join(total_by_month, by = "date") %>%
  mutate(share = imports / total_imports * 100) %>%
  filter(date >= "2024-01-01", date <= Sys.Date())

transship_colors <- c("China" = "#C8102E", "Vietnam" = "#28A745",
                      "Thailand" = "#FFC107", "Malaysia" = "#0055A4",
                      "Cambodia" = "#9B59B6")

p_level <- transship_data %>%
  ggplot(aes(x = date, y = imports_B, color = country)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = transship_colors) +
  labs(title = "Import Levels: China vs. SE Asian Transshipment Countries",
       x = NULL, y = "Billions USD", color = NULL)

p_share <- share_data %>%
  ggplot(aes(x = date, y = share, color = country)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = transship_colors) +
  labs(title = "Import Share: China vs. SE Asian Transshipment Countries",
       subtitle = "Share of total U.S. goods imports (%)",
       x = NULL, y = "Share of Total Imports (%)", color = NULL,
       caption = fig_caption("F5"))

p_level / p_share
```

```{r transship-na, eval=!census_available, results='asis'}
cat("*Census trade data not available.*")
```

---

## Chapter-Level Effective Rates (Census Calculated Duty)

The Census Bureau reports two duty-related fields at the chapter × country × month level:

- **`CAL_DUT_MO`** (Calculated Duty): The estimated duty amount based on tariff schedules and reported values. Per Census: *"Estimates of calculated duty do not necessarily reflect amounts of duty paid."* This overstates actual collections because it doesn't fully account for preference programs.
- **`DUT_VAL_MO`** (Dutiable Value): The value of imports *subject to duty*---not the duty itself. This tells us what share of imports are dutiable vs. duty-free.

The figure below uses dutiable value as a share of consumption value, which indicates the fraction of imports in each chapter that face some tariff. A re-pull with `CAL_DUT_MO` (actual calculated duty amounts) is pending.

```{r fig-chapter-etr, eval=census_available}
# Top 10 chapters by 2025 import value
top_chapters <- census %>%
  filter(year == 2025) %>%
  group_by(hs2) %>%
  summarize(total_imports = sum(con_val_mo, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_imports)) %>%
  head(10) %>%
  pull(hs2)

census %>%
  filter(hs2 %in% top_chapters, year %in% c(2024, 2025)) %>%
  group_by(date, hs2) %>%
  summarize(
    imports = sum(con_val_mo, na.rm = TRUE),
    duties = sum(dut_val_mo, na.rm = TRUE),
    etr = duties / imports * 100,
    .groups = "drop"
  ) %>%
  filter(is.finite(etr), etr < 200) %>%  # Filter extreme outliers
  ggplot(aes(x = date, y = etr, color = hs2)) +
  geom_line(linewidth = 0.6) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  labs(
    title = fig_meta("F6", "title"),
    subtitle = fig_meta("F6", "subtitle"),
    x = NULL, y = fig_meta("F6", "y_label"),
    color = "HS Chapter",
    caption = fig_caption("F6")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1))
```

```{r tbl-chapter-etr, eval=census_available}
# Summary table: 2024 vs 2025 calculated ETR by chapter
chapter_etr_comparison <- census %>%
  filter(year %in% c(2024, 2025)) %>%
  group_by(year, hs2) %>%
  summarize(
    imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9,
    duties_B = sum(dut_val_mo, na.rm = TRUE) / 1e9,
    calc_etr = duties_B / imports_B * 100,
    .groups = "drop"
  ) %>%
  filter(hs2 %in% top_chapters) %>%
  select(year, hs2, imports_B, calc_etr) %>%
  pivot_wider(names_from = year, values_from = c(imports_B, calc_etr),
              names_sep = "_")

chapter_etr_comparison %>%
  mutate(
    `HS Chapter` = hs2,
    `2024 Imports ($B)` = sprintf("%.1f", imports_B_2024),
    `2025 Imports ($B)` = sprintf("%.1f", imports_B_2025),
    `2024 Calc. ETR (%)` = sprintf("%.1f", calc_etr_2024),
    `2025 Calc. ETR (%)` = sprintf("%.1f", calc_etr_2025),
    `ETR Change (pp)` = sprintf("%.1f", calc_etr_2025 - calc_etr_2024)
  ) %>%
  select(`HS Chapter`, `2024 Imports ($B)`, `2025 Imports ($B)`,
         `2024 Calc. ETR (%)`, `2025 Calc. ETR (%)`, `ETR Change (pp)`) %>%
  kable(caption = "Census Calculated ETR by Chapter: 2024 vs. 2025 (Top 10 by Import Value)",
        align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Product-Level Statutory Rates

```{r tbl-chapter-statutory, eval=ts_available}
chapter_summary %>%
  head(20) %>%
  mutate(
    `HS Chapter` = hs2,
    `Products` = n_products,
    `Mean Total Rate (%)` = sprintf("%.1f", mean_total * 100),
    `Mean 232 (%)` = sprintf("%.1f", mean_232 * 100),
    `Mean 301 (%)` = sprintf("%.1f", mean_301 * 100),
    `Mean IEEPA (%)` = sprintf("%.1f", mean_ieepa * 100)
  ) %>%
  select(`HS Chapter`, Products, `Mean Total Rate (%)`,
         `Mean 232 (%)`, `Mean 301 (%)`, `Mean IEEPA (%)`) %>%
  kable(caption = "Top 20 HS Chapters by Mean Statutory Tariff Rate (Current Policy)",
        align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## USMCA Utilization

```{r tbl-usmca-stats, eval=ts_available}
usmca_stats <- current_rates %>%
  filter(country %in% c("1220", "2010")) %>%
  summarize(
    total_pairs = n(),
    usmca_eligible = sum(usmca_eligible, na.rm = TRUE),
    pct_eligible = mean(usmca_eligible, na.rm = TRUE) * 100,
    mean_rate_eligible = mean(total_rate[usmca_eligible], na.rm = TRUE) * 100,
    mean_rate_ineligible = mean(total_rate[!usmca_eligible], na.rm = TRUE) * 100
  )

tibble(
  Metric = c("Total CA/MX product-country pairs",
             "USMCA-eligible pairs",
             "Percent eligible",
             "Mean total rate (eligible, after USMCA adjustment)",
             "Mean total rate (ineligible)"),
  Value = c(
    format(usmca_stats$total_pairs, big.mark = ","),
    format(usmca_stats$usmca_eligible, big.mark = ","),
    sprintf("%.1f%%", usmca_stats$pct_eligible),
    sprintf("%.1f%%", usmca_stats$mean_rate_eligible),
    sprintf("%.1f%%", usmca_stats$mean_rate_ineligible)
  )
) %>%
  kable(caption = "USMCA Eligibility Statistics (Current Policy, Canada + Mexico)") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The Tariff Rate Tracker handles USMCA with a **utilization-share approach**, not a simple binary exemption. Using Census SPI data (Rate Provision field RP=18), the tracker computes product-level USMCA utilization shares---the fraction of each product's CA/MX imports that actually claimed USMCA preferences. Tariff rates are then scaled accordingly:

$$\text{rate}_{\text{adjusted}} = \text{rate}_{\text{statutory}} \times (1 - \text{usmca\_share})$$

The current configuration uses **2025 Census SPI data**, which shows significantly elevated utilization rates relative to pre-tariff levels (Canada ~67%, Mexico ~68%, vs. 2024 pre-tariff steady-state of ~38% and ~50%). This tariff-induced jump in USMCA claiming makes sense: as tariff rates rose, importers had stronger incentives to file USMCA claims for eligible products.

Because the statutory rates already incorporate product-level USMCA utilization, the remaining USMCA contribution to the statutory-actual gap is **smaller** than a naive binary model would suggest. The residual USMCA effect on the gap comes from:

- **Within-product utilization heterogeneity**: The share is an average; some shipments of an eligible product claim USMCA while others don't, creating variance around the mean that doesn't fully cancel.
- **Section 232 auto content scaling**: USMCA-eligible vehicles are only partially exempt, scaled by `us_auto_content_share = 0.40` (reflecting that ~40% of vehicle value is US/USMCA-origin).
- **Timing mismatch**: 2025 utilization rates are applied uniformly, but actual utilization increased over the year as firms adapted.

If anything, the USMCA adjustment may cause the statutory rate to **understate** the true tariff burden for CA/MX, since some of the ~67-68% claiming rate reflects opportunistic filing on products where rules-of-origin compliance is marginal.

---

## Framework for Decomposing the Gap

```{r fig-decomposition, fig.height=6}
decomp <- tibble(
  factor = c("Collection timing lag",
             "USMCA residual (post-share adjustment)",
             "Stockpiling / front-running",
             "Other preferences (GSP, FTAs, de minimis)",
             "Evasion / misclassification",
             "Residual"),
  contribution_pp = c(2.5, -0.3, 1.5, 0.8, 1.0, 0.5),
  direction = ifelse(c(2.5, -0.3, 1.5, 0.8, 1.0, 0.5) >= 0, "Widens gap", "Narrows gap")
)

decomp %>%
  mutate(factor = reorder(factor, contribution_pp)) %>%
  ggplot(aes(x = factor, y = contribution_pp, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Widens gap" = color_gap, "Narrows gap" = color_actual)) +
  geom_hline(yintercept = 0, color = "black") +
  labs(
    title = fig_meta("F7", "title"),
    subtitle = fig_meta("F7", "subtitle"),
    x = NULL, y = fig_meta("F7", "y_label"),
    fill = NULL,
    caption = fig_caption("F7")
  )
```

*Note: The decomposition values shown above are illustrative placeholders. Calibration using the Census chapter-level data is a next step.*

---

## Next Steps {#next-steps}

1. **Calibrate Census calculated duty**: Investigate why `DUT_VAL_MO` diverges so far from actual collections (specific rate treatment, preference pre-application, HMT/MPF inclusion).

2. **Country-level gap analysis**: Using HS2 × country Census data, compute expected vs. actual duty by partner to attribute the gap.

3. **Dynamic stockpiling model**: Estimate the timing of inventory drawdown after each tariff escalation step using a distributed lag model on monthly trade by chapter.

4. **USMCA utilization calibration**: Obtain actual USMCA claim rates from CBP data to refine the USMCA contribution estimate.

5. **Transshipment quantification**: Estimate the share of SE Asian import growth attributable to China trade diversion vs. organic growth.

---

*Report generated on `r Sys.time()` using R `r R.version.string`*
