---
title: "Actual vs. Statutory Effective Tariff Rates, 2025--2026"
author: "The Budget Lab at Yale"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
  word_document:
    toc: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)

here::i_am("R/etr_eval_report.Rmd")
source(here("R", "utils.R"))
theme_set(theme_etr())

# ==============================================================================
# DATA LOADING
# ==============================================================================

# Actual ETR (customs duties / imports, monthly from Census/Treasury via Haver)
actual_etr <- load_actual_etr() %>%
  filter(date >= "2024-01-01") %>%
  mutate(actual_rate = effective_rate / 100)  # convert % to proportion

# Daily statutory ETR (import-weighted, from tariff-rate-tracker)
daily_statutory <- load_daily_etr() %>%
  filter(date <= Sys.Date())

# Authority decomposition
daily_auth <- load_daily_authority()

# Monthly statutory (first-of-month snapshot)
monthly_statutory <- collapse_statutory_monthly(daily_statutory)

# Revision dates for policy event context
revision_dates <- load_revision_dates()

# ==============================================================================
# FIGURE/TABLE METADATA
# ==============================================================================

FIGURE_METADATA <- list(
  F1 = list(
    label    = "Figure 1",
    title    = "Actual vs. Statutory Effective Tariff Rate",
    subtitle = "Monthly actual (customs duties / imports) vs. daily statutory (import-weighted)",
    source   = "U.S. Treasury/Census via Haver Analytics; The Budget Lab Tariff Rate Tracker",
    notes    = "Actual rate is customs duties as a share of goods import value (monthly). Statutory rate is the import-weighted average of all HTS tariff schedules by calendar day.",
    y_label  = "Percent"
  ),
  F2 = list(
    label    = "Figure 2",
    title    = "Statutory-Actual ETR Gap",
    subtitle = "Percentage point difference (statutory minus actual) by month",
    source   = "The Budget Lab analysis",
    notes    = "Positive values indicate the statutory rate exceeds what is actually collected. Gap reflects avoidance, evasion, timing lags, USMCA utilization, and other factors.",
    y_label  = "Percentage Points"
  ),
  F3 = list(
    label    = "Figure 3",
    title    = "Statutory ETR by Trading Partner",
    subtitle = "Import-weighted statutory rate, first-of-month snapshot",
    source   = "The Budget Lab Tariff Rate Tracker",
    notes    = "",
    y_label  = "Percent"
  ),
  F4 = list(
    label    = "Figure 4",
    title    = "Import Volume Shifts and the ETR Gap",
    subtitle = "Monthly real import growth vs. statutory-actual gap, by partner",
    source   = "Census Bureau via Haver Analytics; The Budget Lab analysis",
    notes    = "Stockpiling hypothesis: countries with larger import surges before tariff effective dates may show larger subsequent gaps.",
    y_label  = ""
  ),
  F5 = list(
    label    = "Figure 5",
    title    = "Tariff Gap Decomposition: Potential Explanatory Factors",
    subtitle = "Estimated contribution of avoidance channels to the statutory-actual gap",
    source   = "The Budget Lab analysis",
    notes    = "Decomposition is illustrative; channels are not fully separable from aggregate data alone.",
    y_label  = "Percentage Points"
  )
)

fig_meta <- function(id, field) {
  val <- FIGURE_METADATA[[id]][[field]]
  if (is.null(val)) "" else val
}
fig_caption <- function(id) {
  src <- fig_meta(id, "source")
  notes <- fig_meta(id, "notes")
  cap <- paste0("Source: ", src)
  if (nzchar(notes)) cap <- paste0(cap, "\nNote: ", notes)
  cap
}

# ==============================================================================
# COMPUTED OBJECTS
# ==============================================================================

# Merge actual and statutory at monthly level
comparison <- monthly_statutory %>%
  select(date, statutory_rate = weighted_etr) %>%
  full_join(
    actual_etr %>% select(date, actual_rate),
    by = "date"
  ) %>%
  filter(date >= "2025-01-01") %>%
  arrange(date) %>%
  mutate(
    gap_pp = (statutory_rate - actual_rate) * 100,
    statutory_pct = statutory_rate * 100,
    actual_pct = actual_rate * 100
  )

# Latest comparison stats
latest_comp <- comparison %>% filter(!is.na(actual_rate)) %>% filter(date == max(date))
latest_month <- format(latest_comp$date, "%B %Y")
latest_statutory <- sprintf("%.1f", latest_comp$statutory_pct)
latest_actual <- sprintf("%.1f", latest_comp$actual_pct)
latest_gap <- sprintf("%.1f", latest_comp$gap_pp)
```

## Summary

This report compares two measures of tariff rates on U.S. imports during the 2025--2026 tariff escalation:

- **Statutory ETR**: The import-weighted average tariff rate implied by the Harmonized Tariff Schedule, computed from The Budget Lab's [Tariff Rate Tracker](https://budgetlab.yale.edu/research/state-tariffs). This represents what importers *should* pay given the tariff schedule in effect.
- **Actual ETR**: Customs duties actually collected divided by the value of goods imports, from Treasury/Census data via Haver Analytics. This represents what importers *actually* paid.

As of `r latest_month`, the statutory ETR was `r latest_statutory`% while the actual collected rate was `r latest_actual`%, a gap of `r latest_gap` percentage points. This gap has several potential explanations---timing lags between tariff announcements and collection, preferential trade agreement utilization (USMCA), tariff avoidance through supply chain restructuring, stockpiling ahead of tariff effective dates, and outright evasion---which we explore below.

---

## Figure 1: Actual vs. Statutory ETR

The statutory rate (navy line) represents the daily import-weighted average tariff rate from the HTS schedule. The actual rate (red dots) shows monthly customs duties as a share of import value. The actual rate is inherently lagged: tariffs announced in month $t$ may not appear in collection data until months $t+1$ or $t+2$ due to processing times, bonding, and liquidation schedules.

```{r fig-actual-vs-statutory}
ggplot() +
  # Statutory: daily step line
  geom_step(data = daily_statutory,
            aes(x = date, y = weighted_etr * 100),
            color = color_statutory, linewidth = 0.7) +
  # Actual: monthly scatter
  geom_point(data = actual_etr %>% filter(date >= "2025-01-01"),
             aes(x = date, y = actual_rate * 100),
             color = color_actual, size = 3, shape = 16) +
  # Policy event lines
  geom_vline(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
             aes(xintercept = date), linetype = "dashed",
             color = tbl_gray, linewidth = 0.3) +
  geom_text(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
            aes(x = date, y = Inf, label = label),
            vjust = 1.3, hjust = -0.05, size = 2.3, color = tbl_gray) +
  # Legend annotations
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.95,
           label = "Statutory (daily)", color = color_statutory, hjust = 0, fontface = "bold") +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.88,
           label = "Actual (monthly)", color = color_actual, hjust = 0, fontface = "bold") +
  labs(
    title = fig_meta("F1", "title"),
    subtitle = fig_meta("F1", "subtitle"),
    x = NULL,
    y = fig_meta("F1", "y_label"),
    caption = fig_caption("F1")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_cartesian(ylim = c(0, NA))
```

---

## Figure 2: The Statutory-Actual Gap

```{r fig-gap}
comparison %>%
  filter(!is.na(gap_pp)) %>%
  ggplot(aes(x = date, y = gap_pp)) +
  geom_col(fill = color_gap, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  geom_text(aes(label = sprintf("%.1f pp", gap_pp)),
            vjust = ifelse(comparison$gap_pp[!is.na(comparison$gap_pp)] >= 0, -0.5, 1.5),
            size = 3) +
  labs(
    title = fig_meta("F2", "title"),
    subtitle = fig_meta("F2", "subtitle"),
    x = NULL,
    y = fig_meta("F2", "y_label"),
    caption = fig_caption("F2")
  )
```

The gap between statutory and actual rates is expected to be positive for several structural reasons:

1. **Collection timing**: Tariffs are assessed at entry but may be collected weeks or months later. Monthly revenue data reflects a blend of tariff schedules in effect at the time of import.
2. **USMCA utilization**: Products qualifying under USMCA enter duty-free, but our statutory calculation uses a binary S/S+ flag from HTS rather than actual utilization rates (median ~50% for CA/MX products).
3. **Duty drawback and refunds**: Some collected duties are later refunded through drawback programs.
4. **Preferential programs**: GSP, FTAs, and other preference programs reduce effective rates below MFN schedules.
5. **De minimis threshold**: Shipments below $800 enter duty-free (though this was reduced for some countries in 2025).

---

## Collapsing Statutory Rates to Monthly

We collapse the daily statutory ETR to monthly using **first-of-month snapshots**. This approach is preferable to monthly averaging because:

- Tariff rates are step functions---they change only at discrete policy revision dates, not gradually. Averaging would create artificial intermediate values that never actually applied.
- Most tariff revisions take effect on or near the 1st of the month, so the first-of-month snapshot captures the prevailing rate for most of the month.
- Census reports monthly import values without within-month timing, so there's no additional precision gained from day-weighting.
- It's transparent and easy to interpret: "what was the tariff schedule at month start?"

```{r tbl-monthly-comparison}
comparison %>%
  filter(!is.na(actual_rate)) %>%
  mutate(
    Month = format(date, "%b %Y"),
    `Statutory ETR (%)` = sprintf("%.2f", statutory_pct),
    `Actual ETR (%)` = sprintf("%.2f", actual_pct),
    `Gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Month, `Statutory ETR (%)`, `Actual ETR (%)`, `Gap (pp)`) %>%
  kable(caption = "Monthly Statutory vs. Actual Effective Tariff Rates",
        align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Country-Level Decomposition

```{r load-country-data, include=FALSE}
# Load statutory ETR by partner from the tariff-rate-tracker
etr_partner_file <- file.path(TRACKER_DIR, "output", "daily", "daily_by_country.csv")
etr_partner_available <- file.exists(etr_partner_file)

if (etr_partner_available) {
  daily_country <- read_csv(etr_partner_file, show_col_types = FALSE) %>%
    mutate(date = as.Date(date))

  # Define major partners for display
  MAJOR_PARTNERS <- c("China", "Canada", "Mexico", "EU", "Japan", "S. Korea", "UK")

  # Collapse to monthly first-of-month
  monthly_country <- daily_country %>%
    filter(day(date) == 1, date >= "2025-01-01") %>%
    mutate(
      partner_group = case_when(
        country_name %in% c("China") ~ "China",
        country_name %in% c("Canada") ~ "Canada",
        country_name %in% c("Mexico") ~ "Mexico",
        country_name %in% c("Japan") ~ "Japan",
        country_name %in% c("Korea, South", "South Korea") ~ "S. Korea",
        country_name %in% c("United Kingdom") ~ "UK",
        # EU detection would need country list; placeholder
        TRUE ~ "ROW"
      )
    )
}
```

```{r fig-country-etr, eval=etr_partner_available}
# Show top trading partners' statutory ETR over time
top_partners <- monthly_country %>%
  filter(partner_group %in% MAJOR_PARTNERS) %>%
  group_by(date, partner_group) %>%
  summarize(
    weighted_etr = weighted.mean(weighted_etr, w = pmax(n_products_present, 1), na.rm = TRUE),
    .groups = "drop"
  )

top_partners %>%
  ggplot(aes(x = date, y = weighted_etr * 100, color = partner_group)) +
  geom_step(linewidth = 0.8) +
  scale_color_manual(values = PARTNER_COLORS) +
  labs(
    title = fig_meta("F3", "title"),
    subtitle = fig_meta("F3", "subtitle"),
    x = NULL,
    y = fig_meta("F3", "y_label"),
    color = NULL,
    caption = fig_caption("F3")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1))
```

```{r country-etr-na, eval=!etr_partner_available, results='asis'}
cat("*Country-level daily ETR data not available.*")
```

A key limitation of the country-level comparison is that we do not have country-level *actual* collection data from Treasury. The actual ETR from Haver is an aggregate across all trading partners. Country-level actual rates could in principle be derived from Census IM145/IM148 microdata (which reports duties collected by HTS line item), but this data has significant lags and is not yet available for the full 2025 period.

**Proposed approach for country-level gap analysis**: Use the product-level timeseries RDS to compute *expected* revenue by country (statutory rate $\times$ 2024 import value), then compare against the aggregate actual revenue. Countries with a larger share of total expected revenue but a smaller share of actual collections would indicate disproportionate avoidance or USMCA utilization. This is the approach we develop in the [product-level analysis](#product-level-analysis) below.

---

## Product-Level Analysis {#product-level-analysis}

```{r load-product-data, include=FALSE}
# Load the full timeseries for product-level analysis
# This is a large object (~50M+ rows across revisions); we load it once and filter
ts_available <- file.exists(file.path(TRACKER_DIR, "data", "timeseries", "rate_timeseries.rds"))
if (ts_available) {
  ts <- load_timeseries()

  # Get the latest snapshot
  latest_date <- min(Sys.Date(), max(ts$valid_until, na.rm = TRUE))
  current_rates <- get_rates_at_date(ts, latest_date)

  # Summarize by HS2 chapter
  chapter_summary <- current_rates %>%
    mutate(hs2 = substr(hts10, 1, 2)) %>%
    group_by(hs2) %>%
    summarize(
      n_products = n_distinct(hts10),
      mean_total = mean(total_rate, na.rm = TRUE),
      mean_232 = mean(rate_232, na.rm = TRUE),
      mean_301 = mean(rate_301, na.rm = TRUE),
      mean_ieepa = mean(rate_ieepa_recip, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_total))
}
```

```{r tbl-chapter-rates, eval=ts_available}
chapter_summary %>%
  head(20) %>%
  mutate(
    `HS Chapter` = hs2,
    `Products` = n_products,
    `Mean Total Rate (%)` = sprintf("%.1f", mean_total * 100),
    `Mean 232 (%)` = sprintf("%.1f", mean_232 * 100),
    `Mean 301 (%)` = sprintf("%.1f", mean_301 * 100),
    `Mean IEEPA (%)` = sprintf("%.1f", mean_ieepa * 100)
  ) %>%
  select(`HS Chapter`, Products, `Mean Total Rate (%)`,
         `Mean 232 (%)`, `Mean 301 (%)`, `Mean IEEPA (%)`) %>%
  kable(caption = "Top 20 HS Chapters by Mean Statutory Tariff Rate (Current Policy)",
        align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Explaining the Gap: Avoidance, Evasion, and Stockpiling

The statutory-actual gap reflects a mix of legitimate and illegitimate factors. We consider three main channels:

### 1. Stockpiling / Front-Running

Importers who anticipate tariff increases have strong incentives to accelerate shipments before effective dates. This creates:

- A **pre-tariff import surge**: imports spike in the weeks before a tariff takes effect, entering at the lower (or zero) pre-tariff rate
- A **post-tariff import dip**: imports fall as inventories are drawn down
- A **persistent gap**: during the drawdown period, actual collections lag statutory rates because the goods entering consumption were imported (and taxed) under the old regime

We can test this by examining whether HS chapters that saw the largest tariff increases also experienced the most pronounced import timing shifts. The tariff-impact-tracker's trade deviation analysis (Figures 14--15) documents the aggregate pattern of front-running; here we extend it to the product level.

```{r fig-stockpiling-concept, eval=ts_available, fig.height=7}
# Conceptual figure: compare rate changes vs import timing by chapter
# For each HS2, compute (a) the tariff increase from basic to current,
# (b) how much of total imports that chapter represents

baseline_rates <- get_rates_at_date(ts, "2025-01-01")
baseline_by_ch <- baseline_rates %>%
  mutate(hs2 = substr(hts10, 1, 2)) %>%
  group_by(hs2) %>%
  summarize(baseline_rate = mean(total_rate, na.rm = TRUE), .groups = "drop")

rate_change <- chapter_summary %>%
  left_join(baseline_by_ch, by = "hs2") %>%
  mutate(
    rate_increase = mean_total - baseline_rate,
    rate_increase_pp = rate_increase * 100
  ) %>%
  filter(rate_increase_pp > 1)  # Focus on chapters with meaningful increases

p1 <- rate_change %>%
  ggplot(aes(x = reorder(hs2, rate_increase_pp), y = rate_increase_pp)) +
  geom_col(fill = color_statutory, alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Tariff Increase by HS Chapter",
    subtitle = "Increase in mean statutory rate from Jan 2025 baseline to current policy (pp)",
    x = "HS Chapter",
    y = "Rate Increase (pp)"
  )

p2 <- rate_change %>%
  ggplot(aes(x = rate_increase_pp, y = n_products)) +
  geom_point(color = color_actual, size = 2) +
  geom_text(aes(label = hs2), hjust = -0.2, size = 2.5) +
  labs(
    title = "Rate Increase vs. Product Count",
    subtitle = "Chapters with more products may show more aggregate avoidance",
    x = "Rate Increase (pp)",
    y = "Number of HTS10 Products"
  )

p1 / p2 +
  plot_annotation(
    title = fig_meta("F4", "title"),
    caption = fig_caption("F4")
  )
```

**Proposed extension**: With monthly Census import data by HS chapter (available from Census USA Trade Online or USITC DataWeb), we could compute:

$$\text{Stockpiling Index}_c = \frac{\text{Imports}_{c,\text{pre-tariff}}}{\text{Imports}_{c,\text{same period prior year}}}$$

for each chapter $c$, where the pre-tariff window is the 1--3 months before the tariff effective date. Chapters with a high stockpiling index and a large tariff increase would be expected to show a larger statutory-actual gap in subsequent months as the pre-tariff inventory clears.

### 2. USMCA Utilization

A significant portion of the statutory-actual gap is likely attributable to USMCA. Our statutory rate calculation uses binary S/S+ eligibility flags from the HTS `special` field to identify USMCA-eligible products, but actual USMCA utilization rates are considerably lower---median around 50% for Canadian and Mexican products. This means:

- We assume 100% USMCA utilization for eligible products (rate = 0 on those product-country pairs)
- In reality, only ~50% of eligible trade actually claims USMCA preferences
- The remaining 50% pays the full tariff rate
- This should make statutory rates *lower* than actual for CA/MX products, partially offsetting the gap from other factors

```{r tbl-usmca-stats, eval=ts_available}
# Count USMCA-eligible products
usmca_stats <- current_rates %>%
  filter(country %in% c("1220", "2010")) %>%  # Canada, Mexico census codes
  summarize(
    total_pairs = n(),
    usmca_eligible = sum(usmca_eligible, na.rm = TRUE),
    pct_eligible = mean(usmca_eligible, na.rm = TRUE) * 100,
    mean_rate_eligible = mean(total_rate[usmca_eligible], na.rm = TRUE) * 100,
    mean_rate_ineligible = mean(total_rate[!usmca_eligible], na.rm = TRUE) * 100
  )

tibble(
  Metric = c("Total CA/MX product-country pairs",
             "USMCA-eligible pairs",
             "Percent eligible",
             "Mean total rate (eligible)",
             "Mean total rate (ineligible)"),
  Value = c(
    format(usmca_stats$total_pairs, big.mark = ","),
    format(usmca_stats$usmca_eligible, big.mark = ","),
    sprintf("%.1f%%", usmca_stats$pct_eligible),
    sprintf("%.1f%%", usmca_stats$mean_rate_eligible),
    sprintf("%.1f%%", usmca_stats$mean_rate_ineligible)
  )
) %>%
  kable(caption = "USMCA Eligibility Statistics (Current Policy, Canada + Mexico)") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

### 3. Tariff Evasion and Misclassification

Tariff evasion---intentional misclassification of products to lower-tariff HTS codes, undervaluation of declared import values, or transshipment through third countries---is harder to detect from aggregate data but likely contributes to the gap, particularly for:

- **China-origin goods**: The large IEEPA + 301 surcharges create strong incentives for transshipment through Southeast Asian countries
- **Steel/aluminum derivatives**: Section 232 derivative product identification relies on metal content shares that importers may underreport
- **De minimis exploitation**: Small shipments that avoid duties entirely (though the $800 threshold was tightened in 2025)

**Proposed diagnostic**: Compare the growth in imports from countries commonly used as transshipment points (Vietnam, Thailand, Malaysia, Cambodia) against the decline in Chinese imports. If transshipment is significant, we would expect:

$$\frac{\Delta \text{Imports}_{\text{Vietnam}}}{\Delta \text{Imports}_{\text{China}}} \approx -1$$

for products with high China-specific tariffs (301 + IEEPA).

---

## Framework for Decomposing the Gap

```{r fig-decomposition, fig.height=6}
# Illustrative decomposition (placeholder values -- to be estimated from data)
# The goal is to show the framework; actual estimation requires additional data

decomp <- tibble(
  factor = c("Collection timing lag",
             "USMCA utilization gap",
             "Stockpiling / front-running",
             "Other preferences (GSP, FTAs)",
             "Evasion / misclassification",
             "Residual"),
  contribution_pp = c(2.5, -1.0, 1.5, 0.5, 1.0, 0.5),
  direction = ifelse(c(2.5, -1.0, 1.5, 0.5, 1.0, 0.5) >= 0, "Widens gap", "Narrows gap")
)

decomp %>%
  mutate(factor = reorder(factor, contribution_pp)) %>%
  ggplot(aes(x = factor, y = contribution_pp, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Widens gap" = color_gap, "Narrows gap" = color_actual)) +
  geom_hline(yintercept = 0, color = "black") +
  labs(
    title = fig_meta("F5", "title"),
    subtitle = fig_meta("F5", "subtitle"),
    x = NULL,
    y = fig_meta("F5", "y_label"),
    fill = NULL,
    caption = fig_caption("F5")
  )
```

*Note: The decomposition values shown above are illustrative placeholders. Estimation of each channel requires additional data and analysis described in the [Next Steps](#next-steps) section.*

---

## Next Steps {#next-steps}

1. **Monthly Census trade data by HS chapter**: Pull from Census USA Trade Online or USITC DataWeb to construct product-level stockpiling indices and test the correlation between tariff increases and import timing shifts.

2. **Country-level duties collected**: Investigate IM145/IM148 microdata availability for country-level actual tariff collection, which would allow a direct actual-vs-statutory comparison by partner.

3. **Transshipment analysis**: Decompose the growth in imports from Vietnam, Thailand, Malaysia, and Cambodia by HS chapter and compare against the decline from China.

4. **USMCA utilization rates**: Obtain actual USMCA claim rates from CBP data (published with a lag) to calibrate the USMCA portion of the gap decomposition.

5. **Dynamic stockpiling model**: Estimate the timing of inventory drawdown after each tariff escalation step, potentially using a distributed lag model on monthly trade by chapter.

6. **Reconciliation with tariff-impact-tracker trade figures**: Align the import surge patterns documented in Figures 14--15 of the tariff effects tracker with the product-level gap analysis here.

---

*Report generated on `r Sys.time()` using R `r R.version.string`*
