---
title: "Actual vs. Statutory Effective Tariff Rates, 2025--2026"
author: "The Budget Lab at Yale"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
  word_document:
    toc: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)

here::i_am("R/etr_eval_report.Rmd")
source(here("R", "utils.R"))
theme_set(theme_etr())

# ==============================================================================
# DATA LOADING
# ==============================================================================

# Actual ETR (customs duties / imports, monthly from Census/Treasury via Haver)
actual_etr <- load_actual_etr() %>%
  filter(date >= "2024-01-01") %>%
  mutate(actual_rate = effective_rate / 100)

# Daily statutory ETR (import-weighted, from tariff-rate-tracker)
daily_statutory <- load_daily_etr() %>%
  filter(date <= Sys.Date())

# Monthly statutory (first-of-month snapshot)
monthly_statutory <- collapse_statutory_monthly(daily_statutory)

# Census HS2 x country x month (from 01_pull_census_trade.R)
census_file <- here("data", "census_hs2_country_monthly.csv")
census_available <- file.exists(census_file)
if (census_available) {
  census <- load_census_trade() %>%
    mutate(
      cty_code = as.character(cty_code),
      partner_group = assign_partner_group(cty_code)
    )
}

# Gap decomposition (from 02_decompose_gap.R)
decomp_file <- here("output", "gap_decomposition.csv")
decomp_available <- file.exists(decomp_file)
if (decomp_available) {
  gap_decomp <- read_csv(decomp_file, show_col_types = FALSE) %>%
    mutate(date = as.Date(date))
  gap_by_country <- read_csv(here("output", "gap_by_country.csv"), show_col_types = FALSE)
  gap_by_chapter <- read_csv(here("output", "gap_by_chapter.csv"), show_col_types = FALSE)
}

# Statutory timeseries (product-level)
ts_file <- file.path(TRACKER_DIR, "data", "timeseries", "rate_timeseries.rds")
ts_available <- file.exists(ts_file)
if (ts_available) {
  ts <- load_timeseries()
  latest_date <- min(Sys.Date(), max(ts$valid_until, na.rm = TRUE))
  current_rates <- get_rates_at_date(ts, latest_date)
  baseline_rates <- get_rates_at_date(ts, "2025-01-01")

  chapter_summary <- current_rates %>%
    mutate(hs2 = substr(hts10, 1, 2)) %>%
    group_by(hs2) %>%
    summarize(
      n_products = n_distinct(hts10),
      mean_total = mean(total_rate, na.rm = TRUE),
      mean_232 = mean(rate_232, na.rm = TRUE),
      mean_301 = mean(rate_301, na.rm = TRUE),
      mean_ieepa = mean(rate_ieepa_recip, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_total))

  baseline_by_ch <- baseline_rates %>%
    mutate(hs2 = substr(hts10, 1, 2)) %>%
    group_by(hs2) %>%
    summarize(baseline_rate = mean(total_rate, na.rm = TRUE), .groups = "drop")
}

# ==============================================================================
# FIGURE/TABLE METADATA
# ==============================================================================

FIGURE_METADATA <- list(
  F1 = list(
    label    = "Figure 1",
    title    = "Actual vs. Statutory Effective Tariff Rate",
    subtitle = "Monthly actual (customs duties / imports) vs. daily statutory (import-weighted)",
    source   = "U.S. Treasury/Census via Haver Analytics; The Budget Lab Tariff Rate Tracker",
    notes    = "Actual rate is customs duties as a share of goods import value (monthly). Statutory rate is the import-weighted average of all HTS tariff schedules by calendar day.",
    y_label  = "Percent"
  ),
  F2 = list(
    label    = "Figure 2",
    title    = "Decomposing the Gap: Behavioral vs. Structural",
    subtitle = "Statutory-actual gap split into trade-weight shifts and residual implementation factors",
    source   = "The Budget Lab analysis; Census Bureau International Trade API",
    notes    = "Behavioral = difference between statutory ETR with 2024 weights and actual 2025 weights. Residual = difference between 2025-weighted statutory ETR and Treasury actual collections.",
    y_label  = "Percentage Points"
  ),
  F3 = list(
    label    = "Figure 3",
    title    = "Chapter-Level Behavioral Gap",
    subtitle = "Difference in statutory ETR between 2024 and actual 2025 trade weights, by HS chapter",
    source   = "The Budget Lab analysis",
    notes    = "Positive values indicate 2024 weights overstate the ETR (import mix shifted toward lower-tariff products). Sized by 2024 import value.",
    y_label  = "Behavioral Gap (pp)"
  ),
  F4 = list(
    label    = "Figure 4",
    title    = "Monthly Import Volumes by Trading Partner",
    subtitle = "Billions USD (consumption value), 2024--2025",
    source   = "Census Bureau International Trade API",
    notes    = "",
    y_label  = "Billions USD"
  ),
  F5 = list(
    label    = "Figure 5",
    title    = "Transshipment Indicator: China vs. SE Asia",
    subtitle = "Import levels and shares for China and common transshipment countries",
    source   = "Census Bureau International Trade API",
    notes    = "",
    y_label  = ""
  ),
  F6 = list(
    label    = "Figure 6",
    title    = "Monthly Imports: IEEPA-Affected Chapters Around Liberation Day",
    subtitle = "Top chapters by IEEPA reciprocal rate; note pre-April surge and post-April dip",
    source   = "Census Bureau International Trade API",
    notes    = "",
    y_label  = "Billions USD"
  ),
  F7 = list(
    label    = "Figure 7",
    title    = "Stockpiling Evidence: Q1 Import Surge vs. Tariff Increase",
    subtitle = "Year-over-year Q1 2025 import growth (%) vs. tariff increase from baseline (pp)",
    source   = "Census Bureau International Trade API; The Budget Lab Tariff Rate Tracker",
    notes    = "Each point is an HS2 chapter. Chapters in the upper-right experienced both large tariff increases and large import surges, consistent with front-running.",
    y_label  = "Q1 2025 YoY Import Growth (%)"
  )
)

fig_meta <- function(id, field) {
  val <- FIGURE_METADATA[[id]][[field]]
  if (is.null(val)) "" else val
}
fig_caption <- function(id) {
  src <- fig_meta(id, "source")
  notes <- fig_meta(id, "notes")
  cap <- paste0("Source: ", src)
  if (nzchar(notes)) cap <- paste0(cap, "\nNote: ", notes)
  cap
}

# ==============================================================================
# COMPUTED OBJECTS
# ==============================================================================

comparison <- monthly_statutory %>%
  select(date, statutory_rate = weighted_etr) %>%
  full_join(
    actual_etr %>% select(date, actual_rate),
    by = "date"
  ) %>%
  filter(date >= "2025-01-01") %>%
  arrange(date) %>%
  mutate(
    gap_pp = (statutory_rate - actual_rate) * 100,
    statutory_pct = statutory_rate * 100,
    actual_pct = actual_rate * 100
  )

latest_comp <- comparison %>% filter(!is.na(actual_rate)) %>% filter(date == max(date))
latest_month <- format(latest_comp$date, "%B %Y")
latest_statutory <- sprintf("%.1f", latest_comp$statutory_pct)
latest_actual <- sprintf("%.1f", latest_comp$actual_pct)
latest_gap <- sprintf("%.1f", latest_comp$gap_pp)
```

## 1. Introduction and Headline Comparison

This report compares the **statutory effective tariff rate** (ETR)---the import-weighted average tariff implied by the Harmonized Tariff Schedule---against the **actual effective tariff rate**---customs duties collected divided by goods import value. As of `r latest_month`, the statutory ETR was `r latest_statutory`% while the actual collected rate was `r latest_actual`%, a gap of `r latest_gap` percentage points. Understanding why these measures diverge is critical for evaluating the fiscal and economic impact of the 2025--2026 tariff escalation.

```{r fig-actual-vs-statutory}
ggplot() +
  geom_step(data = daily_statutory,
            aes(x = date, y = weighted_etr * 100),
            color = color_statutory, linewidth = 0.7) +
  geom_point(data = actual_etr %>% filter(date >= "2025-01-01"),
             aes(x = date, y = actual_rate * 100),
             color = color_actual, size = 3, shape = 16) +
  geom_vline(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
             aes(xintercept = date), linetype = "dashed",
             color = tbl_gray, linewidth = 0.3) +
  geom_text(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
            aes(x = date, y = Inf, label = label),
            vjust = 1.3, hjust = -0.05, size = 2.3, color = tbl_gray) +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.95,
           label = "Statutory (daily)", color = color_statutory, hjust = 0, fontface = "bold") +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.88,
           label = "Actual (monthly)", color = color_actual, hjust = 0, fontface = "bold") +
  labs(
    title = fig_meta("F1", "title"),
    subtitle = fig_meta("F1", "subtitle"),
    x = NULL, y = fig_meta("F1", "y_label"),
    caption = fig_caption("F1")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_cartesian(ylim = c(0, NA))
```

```{r tbl-monthly-comparison}
comparison %>%
  filter(!is.na(actual_rate)) %>%
  mutate(
    Month = format(date, "%b %Y"),
    `Statutory ETR (%)` = sprintf("%.2f", statutory_pct),
    `Actual ETR (%)` = sprintf("%.2f", actual_pct),
    `Gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Month, `Statutory ETR (%)`, `Actual ETR (%)`, `Gap (pp)`) %>%
  kable(caption = "Monthly Statutory vs. Actual Effective Tariff Rates",
        align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

We collapse the daily statutory ETR to monthly using **first-of-month snapshots**, preserving the step-function nature of tariff policy. Census reports monthly import values without within-month timing, so day-weighting would add no precision.

---

## 2. Why Do Statutory and Actual Rates Diverge?

A growing literature has examined why predicted or statutory tariff rates diverge from actual collections. We draw on three key studies to frame the analysis.

### Azzimonti (Richmond Fed, August 2025)

[Azzimonti (2025)](https://www.richmondfed.org/publications/research/economic_brief/2025/eb_25-29) provides the most detailed decomposition to date, analyzing the 8.8 percentage point gap between the predicted average ETR (17.5%) and the actual ETR (8.7%) in May 2025. She identifies three channels:

- **Within-country product mix changes** (0.6 pp): Importers reduced volumes of highly tariffed goods or shifted purchase timing. A CFO Survey found 24.6% of respondents planned to "move up purchases."
- **Cross-country import share shifts** (2.7 pp): China imports fell 45% relative to 2024; Vietnam's share rose from 4% to over 6%.
- **Implementation frictions** (4.2 pp, dominant): Shipping timing delays (~30 days from China), deferred ACH payments (6-week lag), legacy exemptions not captured in policy data, customs system lags, and court challenges to IEEPA authority. Azzimonti finds that 31% of Canadian product lines with expected tariffs generated zero collections.

Her conclusion: the gap is primarily an **implementation** phenomenon, not a forecasting failure, and she projects convergence to 10--14% as enforcement catches up.

### Penn Wharton Budget Model (February 2026)

[Penn Wharton (2026)](https://budgetmodel.wharton.upenn.edu/issues/2026/2/23/effective-tariff-rates-and-revenues-updated-february-23-2026) distinguishes "pre-substitution" (holding import patterns fixed) from "post-substitution" (actual) ETRs. By December 2025, the actual ETR was 9.8%. They highlight two key behavioral responses:

- **USMCA surged to 88.2% utilization** by December 2025, up from ~35--50% before tariffs. Canadian and Mexican imports claiming USMCA faced ETRs under 5%.
- **Purchasing pattern changes** reduced tariff revenue by an estimated **$45.7 billion** relative to what a static import composition would have generated.

### Cui, Feng, Lu, and Tomlin (2025)

[Cui et al. (2025)](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=5110124) model the firm-level decision between stockpiling, sourcing diversification, and responsive pricing as tariff risk mitigation strategies. They find that **stockpiling dominates in the short run** when tariff timing is known, but **sourcing diversification is more sustainable** as tariff duration increases.

### Taxonomy of Channels

Drawing on this literature, we organize the sources of the gap into three categories:

| Category | Channel | Direction | Detectable at finer granularity? |
|----------|---------|-----------|----------------------------------|
| **Behavioral** | Trade diversion across countries | Widens gap | Yes: 2024→2025 weight shift |
| | Product mix shifts within countries | Widens gap | Yes: chapter-level reweighting |
| | Stockpiling / front-running | Widens gap (temporary) | Partially: import timing |
| **Structural** | USMCA & preference programs | Depends on model | Small residual (tracker uses utilization shares) |
| | Legacy exemptions & exclusions | Widens gap | No: persists at all levels |
| | De minimis threshold | Widens gap | No: not in HTS schedule |
| **Implementation** | Collection timing & ACH deferral | Widens gap (temporary) | No: aggregate revenue timing |
| | Customs system lags | Widens gap (temporary) | No: persists at all levels |
| | Court challenges (IEEPA) | Widens gap | No: systemic |

The critical test: if the gap is behavioral, it should **shrink** when we reweight from 2024 to actual 2025 trade patterns, because the gap comes from applying 2024 import composition to 2025 rates. If the gap is structural or implementation-driven, it should **persist** at all granularity levels.

---

## 3. Decomposing the Gap: Behavioral vs. Structural

```{r load-decomp, include=FALSE}
# Prep decomposition for plotting
if (decomp_available) {
  decomp_plot <- gap_decomp %>%
    filter(!is.na(etr_treasury)) %>%
    mutate(
      behavioral_pp = gap_behavioral * 100,
      residual_pp = gap_residual * 100,
      total_pp = gap_total * 100,
      behavioral_share = behavioral_pp / total_pp * 100
    )
}
```

We compute the statutory ETR two ways for each month: (a) weighted by **2024 annual trade patterns** (what the rate tracker reports), and (b) weighted by **actual 2025 monthly imports** from Census data. The difference is the behavioral component---how much of the gap is explained by importers shifting *what* and *where* they buy. The residual, after subtracting both the behavioral component and the Treasury actual rate, captures structural and implementation factors.

```{r fig-decomp-stacked, eval=decomp_available, fig.height=6}
decomp_long <- decomp_plot %>%
  select(date, month, behavioral_pp, residual_pp) %>%
  pivot_longer(c(behavioral_pp, residual_pp),
               names_to = "component", values_to = "pp") %>%
  mutate(component = factor(component,
    levels = c("residual_pp", "behavioral_pp"),
    labels = c("Structural / Implementation", "Behavioral (trade weight shift)")))

decomp_long %>%
  ggplot(aes(x = date, y = pp, fill = component)) +
  geom_col(position = "stack", width = 25) +
  geom_text(data = decomp_plot,
            aes(x = date, y = total_pp + 0.5,
                label = sprintf("%.1f pp\n(%0.f%% behavioral)", total_pp, behavioral_share)),
            inherit.aes = FALSE, size = 2.8) +
  scale_fill_manual(values = c(
    "Structural / Implementation" = color_actual,
    "Behavioral (trade weight shift)" = color_statutory
  )) +
  labs(
    title = fig_meta("F2", "title"),
    subtitle = fig_meta("F2", "subtitle"),
    x = NULL, y = fig_meta("F2", "y_label"),
    fill = NULL,
    caption = fig_caption("F2")
  ) +
  theme(legend.position = "bottom")
```

```{r tbl-decomp, eval=decomp_available}
decomp_plot %>%
  mutate(
    Month = month,
    `Statutory ETR, 2024 weights (%)` = sprintf("%.1f", etr_statutory_2024w * 100),
    `Statutory ETR, actual weights (%)` = sprintf("%.1f", etr_statutory_actualw * 100),
    `Treasury actual ETR (%)` = sprintf("%.1f", etr_treasury * 100),
    `Total gap (pp)` = sprintf("%.1f", total_pp),
    `Behavioral (pp)` = sprintf("%.1f", behavioral_pp),
    `Residual (pp)` = sprintf("%.1f", residual_pp),
    `Behavioral share` = sprintf("%.0f%%", behavioral_share)
  ) %>%
  select(Month, `Statutory ETR, 2024 weights (%)`, `Statutory ETR, actual weights (%)`,
         `Treasury actual ETR (%)`, `Total gap (pp)`, `Behavioral (pp)`,
         `Residual (pp)`, `Behavioral share`) %>%
  kable(caption = "Decomposition of Statutory-Actual ETR Gap",
        align = c("l", rep("r", 7))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r decomp-na, eval=!decomp_available, results='asis'}
cat("*Gap decomposition not available. Run R/02_decompose_gap.R first.*")
```

**Key finding**: The behavioral component---the portion of the gap explained by importers shifting their trade patterns away from 2024 baselines---accounts for only **1--3 percentage points** (11--26% of the total gap). The remaining **7--12 percentage points** persists even when statutory rates are weighted by actual 2025 trade flows. This is consistent with Azzimonti's finding that implementation frictions, not forecasting error or behavioral response, drive the bulk of the divergence.

---

## 4. What Drives the Residual?

The large structural/implementation residual (74--89% of the gap) reflects several factors that are not resolved by reweighting:

### Collection Timing and Deferred Payment

Tariffs are assessed at the time of entry but may not appear in Treasury revenue data for weeks or months. Shipping from China takes ~30 days, and U.S. Customs allows ACH-enrolled importers to defer payment up to six weeks. A tariff announced on April 2 (Liberation Day) may therefore not generate revenue until June. This creates a **temporary** gap that should close over time---but the monthly data show the residual remaining large through December 2025, suggesting timing alone is insufficient.

### USMCA and Preference Programs

The Budget Lab's Tariff Rate Tracker uses **product-level USMCA utilization shares** from Census SPI data (Rate Provision RP=18). In 2025, utilization surged to ~67% for Canada and ~68% for Mexico, up from pre-tariff levels of ~38% and ~50%. Because the statutory rate already incorporates these shares, the residual USMCA contribution to the gap is small. However, Penn Wharton reports aggregate USMCA claiming reached 88% by December 2025---higher than the product-level median---suggesting our model may understate USMCA's effect for high-value products.

### Legacy Exemptions and Enforcement Gaps

Azzimonti documents that **31% of Canadian product lines** with expected tariffs generated zero collections in May 2025, and **26% of all products** had actual tariffs more than 5 pp below predictions despite nonzero collections. This is strikingly uneven: electronics and machinery showed 10--19 pp gaps, while transportation equipment had near-complete enforcement. These exemptions---rooted in legacy customs rules, bond insufficiencies, and administrative discretion---are not captured in the HTS schedule and therefore persist at all granularity levels.

### Chapter-Level Detail

```{r fig-chapter-gap, eval=decomp_available}
# Show behavioral gap by chapter, sized by import value
gap_by_chapter %>%
  mutate(gap_pp = gap_pp) %>%
  filter(imports_2024_B > 1) %>%  # Focus on chapters with > $1B in 2024 imports
  ggplot(aes(x = reorder(hs2, gap_pp), y = gap_pp, size = imports_2024_B)) +
  geom_point(color = color_statutory, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = tbl_gray) +
  coord_flip() +
  scale_size_continuous(range = c(1, 6), labels = dollar_format(prefix = "$", suffix = "B")) +
  labs(
    title = fig_meta("F3", "title"),
    subtitle = fig_meta("F3", "subtitle"),
    x = "HS Chapter", y = fig_meta("F3", "y_label"),
    size = "2024 Imports",
    caption = fig_caption("F3")
  )
```

```{r tbl-country-gap, eval=decomp_available}
gap_by_country %>%
  mutate(
    Partner = partner_group,
    `2024 Imports ($B)` = sprintf("%.1f", imports_2024_B),
    `Dec 2025 Imports ($B)` = sprintf("%.1f", imports_actual_B),
    `Import Change (ann. %)` = sprintf("%.1f", import_change_pct),
    `ETR, 2024 weights (%)` = sprintf("%.1f", etr_2024w * 100),
    `ETR, actual weights (%)` = sprintf("%.1f", etr_actualw * 100),
    `Behavioral gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Partner, `2024 Imports ($B)`, `Dec 2025 Imports ($B)`, `Import Change (ann. %)`,
         `ETR, 2024 weights (%)`, `ETR, actual weights (%)`, `Behavioral gap (pp)`) %>%
  kable(caption = "Country-Level Behavioral Gap (December 2025)",
        align = c("l", rep("r", 6))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The country-level behavioral gaps are **small** (< 1 pp for all major partners), confirming that cross-country trade diversion---while real---is not the primary driver of the statutory-actual divergence.

---

## 5. Import Dynamics

### Partner-Level Trade Flows

```{r fig-partner-imports, eval=census_available}
major <- c("China", "Canada", "Mexico", "EU", "Japan", "S. Korea", "UK")

census %>%
  group_by(date, partner_group) %>%
  summarize(imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9, .groups = "drop") %>%
  filter(date >= "2024-01-01", date <= Sys.Date(), partner_group %in% major) %>%
  ggplot(aes(x = date, y = imports_B, color = partner_group)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = PARTNER_COLORS) +
  labs(
    title = fig_meta("F4", "title"),
    subtitle = fig_meta("F4", "subtitle"),
    x = NULL, y = fig_meta("F4", "y_label"),
    color = NULL,
    caption = fig_caption("F4")
  )
```

### Transshipment: China vs. SE Asia

```{r fig-transshipment, eval=census_available, fig.height=7}
transship_ctys <- c("5700", "5520", "5490", "5570", "5550")
transship_names <- c("5700" = "China", "5520" = "Vietnam", "5490" = "Thailand",
                     "5570" = "Malaysia", "5550" = "Cambodia")

transship_data <- census %>%
  filter(cty_code %in% transship_ctys) %>%
  mutate(country = transship_names[cty_code]) %>%
  group_by(date, country) %>%
  summarize(imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9, .groups = "drop") %>%
  filter(date >= "2024-01-01", date <= Sys.Date())

total_by_month <- census %>%
  group_by(date) %>%
  summarize(total_imports = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

share_data <- census %>%
  filter(cty_code %in% transship_ctys) %>%
  mutate(country = transship_names[cty_code]) %>%
  group_by(date, country) %>%
  summarize(imports = sum(con_val_mo, na.rm = TRUE), .groups = "drop") %>%
  left_join(total_by_month, by = "date") %>%
  mutate(share = imports / total_imports * 100) %>%
  filter(date >= "2024-01-01", date <= Sys.Date())

transship_colors <- c("China" = "#C8102E", "Vietnam" = "#28A745",
                      "Thailand" = "#FFC107", "Malaysia" = "#0055A4",
                      "Cambodia" = "#9B59B6")

p_level <- transship_data %>%
  ggplot(aes(x = date, y = imports_B, color = country)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = transship_colors) +
  labs(title = "Import Levels", x = NULL, y = "Billions USD", color = NULL)

p_share <- share_data %>%
  ggplot(aes(x = date, y = share, color = country)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  scale_color_manual(values = transship_colors) +
  labs(title = "Import Share", x = NULL, y = "Share of Total Imports (%)", color = NULL)

(p_level / p_share) +
  plot_annotation(
    title = fig_meta("F5", "title"),
    caption = fig_caption("F5")
  )
```

### Liberation Day Dynamics

```{r fig-liberation-day, eval=census_available && ts_available}
if (ts_available) {
  ieepa_chapters <- chapter_summary %>%
    filter(mean_ieepa > 0.05) %>%
    arrange(desc(mean_ieepa)) %>%
    head(8) %>%
    pull(hs2)
} else {
  ieepa_chapters <- c("84", "85", "87", "39", "94", "73", "40", "90")
}

census %>%
  filter(hs2 %in% ieepa_chapters, year %in% c(2024, 2025)) %>%
  group_by(date, hs2) %>%
  summarize(imports_B = sum(con_val_mo, na.rm = TRUE) / 1e9, .groups = "drop") %>%
  ggplot(aes(x = date, y = imports_B, color = hs2)) +
  geom_line(linewidth = 0.6) +
  geom_vline(xintercept = as.Date("2025-04-02"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2025-04-02"), y = Inf, label = "Liberation Day",
           vjust = 1.5, hjust = -0.1, size = 3, color = "red") +
  geom_vline(xintercept = as.Date("2025-01-01"), linetype = "dashed", color = tbl_gray) +
  labs(
    title = fig_meta("F6", "title"),
    subtitle = fig_meta("F6", "subtitle"),
    x = NULL, y = fig_meta("F6", "y_label"),
    color = "HS Chapter",
    caption = fig_caption("F6")
  )
```

### Stockpiling: Import Surge vs. Tariff Increase

```{r fig-stockpiling, eval=census_available && ts_available, fig.height=7}
q1_2025 <- census %>%
  filter(year == 2025, month %in% 1:3) %>%
  group_by(hs2) %>%
  summarize(imports_q1_2025 = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

q1_2024 <- census %>%
  filter(year == 2024, month %in% 1:3) %>%
  group_by(hs2) %>%
  summarize(imports_q1_2024 = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

import_growth <- q1_2025 %>%
  inner_join(q1_2024, by = "hs2") %>%
  mutate(yoy_growth = (imports_q1_2025 / imports_q1_2024 - 1) * 100) %>%
  filter(is.finite(yoy_growth))

rate_change <- chapter_summary %>%
  left_join(baseline_by_ch, by = "hs2") %>%
  mutate(rate_increase_pp = (mean_total - baseline_rate) * 100)

stockpile_data <- import_growth %>%
  inner_join(rate_change %>% select(hs2, rate_increase_pp, n_products), by = "hs2") %>%
  mutate(imports_B = imports_q1_2025 / 1e9)

p_scatter <- stockpile_data %>%
  filter(rate_increase_pp > 0.5) %>%
  ggplot(aes(x = rate_increase_pp, y = yoy_growth, size = imports_B)) +
  geom_point(color = color_statutory, alpha = 0.6) +
  geom_text(aes(label = hs2), hjust = -0.3, vjust = 0.5, size = 2.5, color = tbl_gray) +
  geom_hline(yintercept = 0, linetype = "dashed", color = tbl_gray) +
  geom_smooth(method = "lm", se = TRUE, color = color_actual, linewidth = 0.8) +
  scale_size_continuous(range = c(1, 8), labels = dollar_format(prefix = "$", suffix = "B")) +
  labs(
    title = fig_meta("F7", "title"),
    subtitle = fig_meta("F7", "subtitle"),
    x = "Tariff Increase from Jan 2025 Baseline (pp)",
    y = fig_meta("F7", "y_label"),
    size = "Q1 2025 Imports",
    caption = fig_caption("F7")
  )

p_bar <- stockpile_data %>%
  filter(rate_increase_pp > 2) %>%
  arrange(desc(yoy_growth)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(hs2, yoy_growth), y = yoy_growth)) +
  geom_col(fill = color_statutory, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  coord_flip() +
  labs(title = "Top 15 Chapters: Q1 Import Growth (Tariff-Affected Only)",
       subtitle = "Among chapters with > 2pp tariff increase",
       x = "HS Chapter", y = "YoY Import Growth (%)")

p_scatter / p_bar
```

---

## 6. Does the Gap Close Over Time?

```{r fig-gap-over-time, eval=decomp_available}
decomp_plot %>%
  select(date, month, behavioral_pp, residual_pp, total_pp) %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = total_pp, color = "Total gap"), linewidth = 1) +
  geom_line(aes(y = residual_pp, color = "Structural / Implementation"), linewidth = 1) +
  geom_line(aes(y = behavioral_pp, color = "Behavioral"), linewidth = 1) +
  geom_hline(yintercept = 0, color = "black") +
  scale_color_manual(values = c(
    "Total gap" = "black",
    "Structural / Implementation" = color_actual,
    "Behavioral" = color_statutory
  )) +
  labs(
    title = "Statutory-Actual Gap Over Time",
    subtitle = "Does the implementation residual converge as customs systems catch up?",
    x = NULL, y = "Percentage Points", color = NULL,
    caption = "Source: The Budget Lab analysis"
  )
```

Azzimonti (2025) predicted convergence to 10--14% as enforcement aligned with announced policy. The data show **limited convergence**: the residual implementation gap remained at 7--12 pp through December 2025. Several factors may explain the persistence:

- The tariff regime continued changing throughout 2025 (Phase 2 in August, Section 122 in September), meaning customs systems never had time to fully adjust to any single policy.
- Bond insufficiencies reached record levels ($3.6 billion in fiscal 2025), creating systemic collection delays.
- The Supreme Court's February 2026 ruling striking down IEEPA tariffs introduced additional uncertainty, potentially reducing enforcement intensity in late 2025 as legal challenges mounted.

---

## 7. Next Steps and Data Gaps {#next-steps}

1. **Census calculated duty (`CAL_DUT_MO`)**: Re-pull from Census API (pending rate limit reset) to compare product-level statutory duty amounts against Treasury aggregate collections. This provides a third measure---Census's own estimate of what duties should be, computed at the transaction level.

2. **Country-level actual collections**: Investigate IM145/IM148 microdata availability for country-level duty collection, which would allow direct actual-vs-statutory comparison by partner.

3. **USMCA calibration**: Obtain actual USMCA claim rates from CBP data and compare against the Tariff Rate Tracker's product-level utilization shares. Penn Wharton's 88% aggregate figure vs. our 67--68% product-level median suggests heterogeneity worth investigating.

4. **Dynamic stockpiling model**: Estimate inventory drawdown timing after each tariff escalation step using a distributed lag model on monthly chapter-level trade.

5. **Enforcement heterogeneity**: Following Azzimonti's finding of stark industry-level differences, examine whether enforcement gaps correlate with industry lobbying, product complexity, or customs processing capacity.

---

*Report generated on `r Sys.time()` using R `r R.version.string`*
